---
title: "The Effect of Driving Conditions on Crash Severity"
subtitle: "Statistics 302: Accelerated Introduction to Statistics"
author: "Crashcourse: Sihan Li, Rui Pang, Meng Su, and Ellen Wieland"
date: "May 10, 2019"
output: pdf_document
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{Wieland}
- \lhead{Crashcourse}
- \cfoot{\thepage}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(knitr)
crash = read.csv("crash.csv")  
library(Lock5Data)
```

\centering
\raggedright
\newpage
\tableofcontents

\newpage

## Abstract 

By studying the many factors that influence drivers on the road, this observational study analyzed the relationship between driving conditions and crash severity with the intent of furthering the understanding of driving hazards and risks. The data come from the Iowa state Department of Transportation and were collected over ten years. This study looks at the driving conditions of seat belt use, drug impairment, and road conditions to understand their impacts on fatality rates, vehicular damage, and injury severity respectively. The tests found that the use of seat belts is associated with a significantly lower proportion of fatalities than lack of use, and as such this report recommends the use of seat belts when driving, even though a causal relationship cannot be obtained from these data. The data also show that the mean number of vehicles damaged in a crash is significantly lower for crashes involving drugs than crashes that did not, which refutes the claim that impaired driving is more hazardous to others, yet it is never condoned. Finally, the data show that there is a significant association between the road conditions and the severity of the crash for the individuals involved. On the whole, this report looks at some of the conditions that determine a driver's safety and the outcomes of potential crashes in Iowa from 2008 to the present. 


## Introduction 

The development of the seat belt in 1959 and its subsequent legal regulation has helped save hundreds of thousands of lives. In 2016 alone, an estimated 14,668 people in the United States were saved by their seat belts when involved in a car crash (Department of Transportation). In almost every state in the US, failure to wear a seat belt while driving a motor vehicle is punishable by fine. In many others, passengers can also be subjected to fines (GHSA). 

The reason seat belts are held in such high regard by the law is the fact that safety is never guaranteed while on the road. High levels of rainfall or frost can make roads slippery and dangerous. The US Department of Transportation tracks many "road weather variables" such as air temperature, precipitation, fog, pavement temperature, and pavement condition as well as the effects they have on the roadway and traffic flow (FHWA). In a study conducted in the Northeastern United States looking at fatal car crashes on slippery roads, researchers found that the ratios of crashes were highest among drivers aged sixteen to nineteen (Marmor and Marmor). There was also a strong relationship between youth and use of excessive speed for the given road conditions. 

Even the most responsible driver is at risk of mistakes made by other, less cautious drivers, some of whom could be under the influence of drugs or alcohol. Young drivers, especially those driving with their peers, are at a high risk of becoming distracted and being involved in an accident (Ehsani). Drivers may have the right of way and be proceeding through intersections safely, but others may lose control of their vehicle or run a red light, putting everyone at risk. 

This study is set to further understand the myriad conditions that might make roads hazardous and the ways they threaten lives, and potentially present suggestions for improving drivers' safety everywhere. Especially in the United States, driving is a part of everyday life. While many studies have been conducted in large cities or on the coasts, this study is set in the Midwest, specifically Iowa, to look at conditions that many others have not. 
 
The purpose of this observational study is to examine the effects of driving conditions on crash severity and understand which conditions create the most dangerous situations so as to suggest improvements to safety procedures, which are potentially applicable to areas beyond the Midwest. This will be done first by analyzing the relationship between seat belt use and number of fatalities to see how preexisting safety mechanisms in cars protect passengers during accidents. Second, the involvement of drugs or alcohol in the crash and the number of vehicles involved will be studied. This part of the study pushes to understand the extent to which impaired driving puts other drivers at risk. Finally, the relationship between road conditions and the severity of the crash will be analyzed to find which, if any, conditions can be classified as the most dangerous to drivers. 

The primary data comes from a collection of information from the Iowa state Department of Transportation about crashes over a ten year period, which is to be paired with historical weather data for the given timestamp and geographic location to comprehensively understand the environment in which the accidents occurred. In this study, "driving conditions" are understood to mean not only the state of the roads, but the drivers' mentalities as well. 


## Methods 

The data to be analyzed comes from an observational study of 897,413 recorded crash accidents in Iowa from 2008 to 2017, as collected by the Iowa state Department of Transportation (DOT). Since the data comes exclusively from one state, the results of these tests cannot be generalized too far beyond those bounds. The data is representative of all potential car accidents in the state of Iowa between 2008 and the present, and thus that is the population of interest, though similarities may exist in the road conditions and crash risks of many other Midwestern states. 

This is an observational study since no variables were controlled by the DOT; rather, information was collected as events occurred. While this precludes the data from determining causal relationships between driving conditions and crash severity, associations between them can still be found. Running an experimental study around road conditions and car crashes would be extremely unethical, since that would include purposefully placing individuals in harm's way. Thus an observational study is preferred to an experimental study in this instance. 

The data set includes information about the individuals injured in the crash and will aid in the analysis of road conditions and crash severity for all car accidents in the state of Iowa. In addition to this data, historical weather patterns as recorded by the National Oceanic and Atmospheric Association (NOAA) and the National Climatic Data Center (NCDC) will provide measurements of precipitation. The Iowa DOT dataset includes information about the date, time, and location (in longitude and latitude) of all accidents, which can be paired with similar information from the NOAA to yield precise measurements of weather conditions, especially volumes of rain, snow, and sleet that fell preceding the event. 



### Explanatory Variables 

| Variable       |   Type      |  Level                          |
|:---------------|:-----------:|--------------------------------:|
|Road Conditions | Categorical |Wet, Dry, Gravel, Ice/Frost, Sand|
|Precipitation   |Quantitative |                                 |
|Seat Belt Use   | Categorical |Shoulder/Lap Belt, None Used     |
|Age of Driver   |Quantitative |                                 |
|Drug Involvement|Categorical  | Yes, No                         |

### Response Variables 

| Variable           |   Type      |  Level                                            |
|:-------------------|:-----------:|--------------------------------------------------:|
|Severity of Accident| Categorical |Fatal, Minor Injury, Major Injury, Possible/Unknown|
|Number of Fatalities| Quantitative|                                                   |
|Airbag Deployment   | Categorical | Yes, No                                           |
|Number of Vehicles  |Quantitative |                                                   |


Three statistical tests will be run to further analyze the effect of driving conditions on crash severity. First, to analyze the relationship between seat belt use and crash fatalities, a two-sample randomization test of difference of proportions will be conducted. The null hypothesis will be that regardless of seat belt use, the proportion of accidents that result in fatalities are equal. This test will be run for the alternative hypothesis that the use of seat belts are associated with a lower proportion of fatal crashes than not using a seat belt. 

The second test will be a two-sample t-test to analyze the relationship between drug use and number of vehicles involved in the crash to study how impaired driving threatens others on the road. The two groups being studied will be the crashes that did involve alcohol or drugs and the ones that didn't and the mean number of cars involved in crashes for each group will be compared. The null hypothesis is that the mean number of damaged cars does not differ between crashes where drugs were involved and crashes where they were not. The alternative hypothesis will be that crashes involving drugs are associated with a higher mean number of damaged vehicles than crashes without drug involvement. To run the t-test, the samples' means and standard deviations will be utilized to generate a test statistic in the form $t = \frac{\bar{x}_d - \bar{x}_{nd}}{\sqrt{\frac{s_d^2}{n_d} + \frac{s_{nd}^2}{n_{nd}}}}$ and the p-value will be calculated from a t-distribution to determine if the data provide evidence to suggest that impaired driving is associated with a greater mean number of damaged cars at the significance level $\alpha$ = 0.05. 

Finally, a chi-squared test will be conducted to study the relationship between road conditions and severity of the accident. The null hypothesis for this test will state that all road conditions are expected to produce the same number of crashes for each level of severity. The alternative hypothesis will state that the severity of the crash differs for each road condition. This will test to see if the data provides evidence to suggest that some conditions produce more severe accidents than others. 

The analysis will be conducted using RStudio version 1.1.463. Additional R packages to be used will be 'knitr' version 1.21, 'dplyr' version 0.7.8, and 'ggplot2' version 3.1.0, and some of the data comes from the 'rnoaa' package version 0.8.4. 

## Results 


### Difference in Proportions Randomization Test 

#### Original Sample

```{r rand data, echo=FALSE}
Not_belt<- crash$FATALITIES[which(crash$ZINJ_OCCPROTECT == "None used")]

Number_notbelt<-length(which(Not_belt > 0 ))
Number_notbelt1<-length(which(Not_belt ==1 ))
Number_notbelt2<- length(which(Not_belt==2))
Number_notbelt3<- length(which(Not_belt==3))
Number_notbelt4<- length(which(Not_belt==4))
Number_notbelt5<- length(which(Not_belt==5))

Total_Number_notbelt<- Number_notbelt1+Number_notbelt2+Number_notbelt3+
  Number_notbelt4+Number_notbelt5

Total_notbelt_persons<- Number_notbelt1+ 2*Number_notbelt2+ 3*Number_notbelt3+
  4*Number_notbelt4 + 5*Number_notbelt5

Belt<- crash$FATALITIES[which(crash$ZINJ_OCCPROTECT != "None used" | 
                                crash$ZINJ_OCCPROTECT != "Not applicable")]
Number_belt <- length(which(Belt > 0 ))
Number_belt1 <- length(which(Belt == 1))
Number_belt2 <- length(which(Belt == 2))
Number_belt3 <- length(which(Belt == 3))
Number_belt4 <- length(which(Belt == 4))
Number_belt5 <- length(which(Belt == 5))

Total_Number_belt<- Number_belt1 + Number_belt2 + Number_belt3 + Number_belt4 +
  Number_belt5 

Total_belt_persons<- Number_belt1 + 2*Number_belt2 + 3*Number_belt3 + 
  4*Number_belt4 + 5*Number_belt5 

Belted<- length(Belt)
Notbelted<- length(Not_belt)

prop_notbelt<- Total_notbelt_persons/Notbelted
prop_belt<- Total_belt_persons/Belted

```

#### Hypotheses

To test to see if there is a difference between the two proportions, the hypotheses are

$$
\begin{aligned}
  H_0: p_b = p_n \\
  H_a: p_b \neq p_n \\
\end{aligned}
$$

Where $p_b$ is the proportion of crashes including a fatality of an individual wearing a seat belt and $p_n$ is the proportion of crashes with a fatality for a non-belted individual. 


```{r rand randomization, echo=FALSE}
set.seed(302)

X.nb <- Total_notbelt_persons
X.bb <- Total_belt_persons
n.nb <- Notbelted
n.bb <- Belted
p.hat.nb <- X.nb/n.nb
p.hat.bb <- X.bb/n.bb

p.diff <- p.hat.nb-p.hat.bb
X <- X.nb+X.bb
n <- n.nb+n.bb

p.hat <- X/n
B <- 10000

rand.prop.nb <- rbinom(B,n.nb,p.hat)/n.nb
rand.prop.bb <- rbinom(B,n.bb,p.hat)/n.bb
rand.diff <- rand.prop.nb-rand.prop.bb

```

#### Randomization Distribution

```{r rand dist, echo=FALSE, eval=FALSE}
df.rand <- data.frame(rand.diff)
ggplot(df.rand, aes(x=rand.diff)) + 
  geom_density(aes(color=factor(1),fill=factor(1)),alpha=0.1) + 
  labs(y = "Density", 
       title = "Randomization distribution of the difference between the proportions of 
       fatalities associated with and without seat belts") + 
  theme_classic() + 
  theme(plot.title=element_text(size=10),
        axis.title=element_text(size=10),
        legend.title=element_text(size=10)) + 
    theme(legend.position="none")

```

#### Check for Assumptions

Note that the randomization distribution is approximately symmetric and bell-shaped.

#### Compute p-value

```{r rand p, echo=FALSE, eval=FALSE}
tol <- 1.0e-12
mean(abs(rand.diff) >= abs(p.diff-tol))
```

The p-value can be calculated as the proportion of test statistics generated from the randomization distribution that are greater than or equal to the sample test statistic.

#### Interpretation

In this test, the p-value is zero. Thus the data provide evidence to suggest that the difference between the proportion of fatalities associated with crashes where the individuals were wearing seat belts and without seat belts is significantly different. Therefore, there is significant evidence such that difference between the proportion fatalities associated with seat belts and without seat belts is significantly different.(two-tailed two independent sample proportions randomization test, $n_nb=20082$, $n_bb=874582$, $p=0$, $\alpha=0.05$).

### Difference in Means T-Test 

The second test will be a t-test to analyze the relationship between drug use and number of vehicles involved in the crash to study how impaired driving threatens others on the road. The two groups being studied will be the crashes that did involve alcohol or drugs and the ones that didn't and the mean number of cars involved in crashes for each group will be compared.

```{r t data, echo=FALSE}
drug = crash$VEHICLES[which(crash$DRUGALCREL != "None Indicated")]
drug[is.na(drug)]<- 0

no_drug = crash$VEHICLES[which(crash$DRUGALCREL == "None Indicated")]
no_drug[is.na(no_drug)]<- 0

```

#### Hypotheses

To test whether the mean number of vehicles damaged in a crash differs between crashes that involved drugs or alcohol and those that did not, the hypotheses are 

$$
\begin{aligned}
  H_0: \mu_n = \mu_d \\
  H_a: \mu_n < \mu_d \\
\end{aligned}
$$
where $\mu_n$ is the mean number of vehicles damaged in crashes where no drugs were involved and $\mu_d$ is the mean number of vehicles damaged in crashes without any drugs. 

```{r}
summary(drug)
str(drug)
head(drug)
summary(no_drug)
str(no_drug)
head(no_drug)
```

#### Summary Figure and Check for Assumptions

```{r t assumptions, echo=FALSE}
length(drug)
length(no_drug)
boxplot(drug)
boxplot(no_drug)

```

Even though the data show some skewedness, the extremely large samle size still allows for the use of a t-test. 

#### Calculate Test Statistic

$$
\begin{aligned}
  t&=\frac{\bar{x}_n-\bar{x}_d}{\sqrt{\frac{s_d^2}{n_d}+\frac{s_n^2}{n_n}}} \\
  &=\frac{0.67-0.56}{\sqrt{\frac{0.915^2}{41475}+\frac{1.093^2}{855938}}} \\
  &=25.32948 \\
\end{aligned}
$$

```{r t statistic, echo=FALSE, eval=FALSE}
x.bar.d <- mean(drug)
s.d <- sd(drug)
n.d <- length(drug)
x.bar.n <- mean(no_drug)
s.n <- sd(no_drug)
n.n <- length(no_drug)
se <- sqrt((s.n^2/n.n)+(s.d^2/n.d))
t <- (x.bar.n-x.bar.d)/se
t
```

#### Compute p-value

$$
\begin{aligned}
  T&\sim t_{n-1} \\
  p-value&=P(T\geq t) \\
  &=P(T\geq 25.32948) \\
  &=1 \\
\end{aligned}
$$
```{r}
# pt(t, min(n.n-1,n.d-1))
t.test(no_drug,drug,alternative = "less", conf.level = 0.95)
```

The p-value can be calculated as the area to the left of the test statistic on a t distribution with 41,474 degrees of freedom. A t distribution is valid for these data since the sample sizes are very large and do not have outliers. The p-value for this test is one, so there is no significant evidence to suggest that the mean number of vehicles damaged in crashes that did not involve drugs is fewer than the mean number of vehicles damaged in crashes that did involve drugs. 

If the test is run with a different alternative hypothesis, however, such as 

$$
\begin{aligned}
  H_0: \mu_n = \mu_d \\
  H_a: \mu_n > \mu_d \\
\end{aligned}
$$

the same test can be run with the same test statistic, but this time the p-value is zero and the data reject the null hypothesis in favor of the alternative. This shows that crashes that involve drugs are associated with a lower mean number of damaged cars than crashes that did not involve drugs, which goes against initial thinking but is statistically significant. 

#### Interpretation

There is significant evidence such that the mean vehicles involved that crashed by people who did not have drug is equal to the mean vehicles involved that crashed by people who did have had drug. This shows that crashes that involve drugs are associated with a lower mean number of damaged cars than crashes that did not involve drugs, which also against initial thinking. (left-tail two-sample independent t-test,$n_d=41475$, $n_n=855938$, $p=1$, $\alpha=0.05$)


### Chi-Squared Test of Association 

#### Hypotheses

To test for association between road conditions and crash severity, a $\chi^2$ test for independence will be run with the hypotheses: 

$$
\begin{aligned}
  H_0: \texttt{Road Condition and Crash Severity are not associated.}  \\
  H_a: \texttt{Road Condition is associated with Crash Severity.} \\
\end{aligned}
$$

#### Summary of Data

```{r, echo=FALSE}
csuf<- crash$CSURFCOND
cse<-  crash$CSEVERITY
summary(csuf)
str(csuf)
head(csuf)
summary(cse)
str(cse)
head(cse)
obs.counts <- table(crash$CSURFCOND, crash$CSEVERITY)
kable(table(crash$CSURFCOND, crash$CSEVERITY), align="l")
r <- nrow(obs.counts)
c <- ncol(obs.counts)

row.sums <- rowSums(obs.counts)
col.sums <- colSums(obs.counts)
n <- sum(obs.counts)

```

#### Summary Figure

```{r, echo=FALSE}
crash %>% 
  ggplot(aes(x=CSURFCOND,fill=CSEVERITY)) +
  geom_bar(aes(color=CSEVERITY),position="dodge",alpha = 0.25) + 
  labs(x = "Condition of road", 
       y = "Count", 
       title = "Crash severity",
       fill = "Top crash",
       color = "Top crash") + 
  theme_classic() + 
  theme(plot.title=element_text(size=1),
        axis.title=element_text(size=1),
        legend.title=element_text(size=1))

```

#### Check for Assumptions

```{r, echo=FALSE}
exp.counts <- outer(row.sums,col.sums,"*")/n
exp.counts>=5
prod(exp.counts>=5)

chisq.test(obs.counts,correct=FALSE)
```

Some of the expected counts are less than 4. Therefore, it is not appropriate to use the $\chi^2$-distribution.

```{r, echo=FALSE}
crash1 <- crash %>% 
  filter(CSURFCOND != "Oil") %>% 
  mutate(CSURFCOND = factor(CSURFCOND))

crash2 <- crash1 %>% 
  filter(CSURFCOND != "Sand") %>% 
  mutate(CSURFCOND = factor(CSURFCOND))

crash3 <- crash2 %>% 
  filter(CSURFCOND != "Other") %>%
  mutate(CSURFCOND = factor(CSURFCOND))

crash4 <- crash3 %>% 
  filter(CSURFCOND != "Water (standing or moving)") %>% 
  mutate(CSURFCOND = factor(CSURFCOND))

new.counts <- table(crash4$CSURFCOND, crash4$CSEVERITY)

r <- nrow(new.counts)
c <- ncol(new.counts)

row.sums <- rowSums(new.counts)
col.sums <- colSums(new.counts)
n <- sum(new.counts)
  
```

#### Summary Figure

```{r, echo=FALSE}
crash4 %>% 
  ggplot(aes(x=CSURFCOND,fill=CSEVERITY)) +
  geom_bar(aes(color=CSEVERITY),position="dodge",alpha = 0.25) + 
  labs(x = "Condition of road", 
       y = "Count", 
       title = "Crash severity",
       fill = "Top crash",
       color = "Top crash") + 
  theme_classic() + 
  theme(plot.title=element_text(size=15),
        axis.title=element_text(size=15),
        legend.title=element_text(size=1))

```

#### Check for Assumptions

```{r, echo=FALSE}
exp.counts <- outer(row.sums,col.sums,"*")/n
exp.counts>=5
prod(exp.counts>=5)
```

All the expected counts are 5 or greater. Therefore, it is appropriate to use the $\chi^2$-distribution.

#### Calculate Test Statistic

```{r, echo=FALSE}
X.sq <- sum((new.counts-exp.counts)^2/exp.counts)
```

$$
\begin{aligned} 
  X^2&=\sum_{i=1}^k\frac{(x_i-np_{0i})^2}{np_{0i}} \\
  \end{aligned}
$$

#### Compute p-value

```{r, echo=FALSE}
pchisq(X.sq,(r-1)*(c-1),lower.tail=FALSE)
1-pchisq(X.sq,(r-1)*(c-1))

chisq.test(new.counts,correct=FALSE)
chisq.test(table(crash4$CSURFCOND, crash4$CSEVERITY),correct=FALSE)

```

$$
\begin{aligned}
  X^2 &\sim \chi^2_{(r-1)(c-1)} \\
  p-value&=P(\chi^2_{(r-1)(c-1)}\geq X^2) \\
  &=P(\chi^2_{40} \geq 34401.77) \\
  &=0 \\
\end{aligned}
$$
#### Interpretation

There is significant evidence such that crash severity is associated with road conditions. (chi-square test for association, $X^2=34401.77$, $df=40$, $p=0$, $\alpha=0.05$).

The $\chi^2$ statistic is 35,567 and on a distribution with 60 degrees of freedom, the corresponding p-value is basically zero. Thus the data give reason to reject the null hypothesis and show that crash severity is associated with road conditions. 

## Discussion 

The purpose of this observational study is to understand the relationship between driving conditions and crash severity. 

This was done first through a difference in proportions randomization test to study whether or not there was a difference in the proportion of accidents that resulted in fatalities for individuals wearing seat belts and individuals who did not. The purpose of this test in the context of the overarching question was to study one aspect of driving conditions that was implemented specifically to save lives during crashes. Thus crash severity was measured in terms of proportion of crashes that resulted in fatalities to study how seat belts work to save lives during car accidents. The test found that the data provided significant evidence to suggest that wearing a seatbelt is associated with fewer fatalities. While this cannot show causation, the data provide some support for the claim that seat belts are an effective safety implementation. 

Then a two sample t-test was implemented to study the relationship between drug use and the mean number of vehicles damaged in the crash. Drug use would impair a driver's function and awareness, and is thus a measure of driving conditions, while the number of vehicles involved in the crash measured the severity of the accident. The initial hypothesis was that drug users posed an undue threat to other drivers on the road, and thus the mean number of damaged vehicles would be greater for crashes involving drugs or alcohol, but the data suggested the opposite to be true. The drug-related crashes were shown to have a significantly lower mean than the crashes without drugs, but the data do not offer any explanations as to why.

Finally, a chi-square test for independence was conducted to test whether or not there was an association between physical road conditions and crash severity. The road conditions measured elements such as the material the road was constructed from and weather elements that could influence the environment that the driver is in. The data set has its own measure of crash severity for each accident, which measured the extent to which a person was injured. The test found that there was strong evidence in the data to show an association between those variables, which opens the door for further analysis of which road conditions associated with the most severe crashes.

#### Error Analysis



#### Further Study 

"Driving conditions" and "crash severity" continue to be broad categories; this observational study analyzes three factors of each and how they relate, but many more studies can be done to better understand the hazards associated with driving. 

Further tests could include analyzing the crashes that occurred in "wet" conditions and the amount of rainfall in the hours preceeding the accident. This could provide more insight into the relationship between inclement weather and drivers' safety. Another test could analyze the age of the driver and factors such as drug use or severity of crash to compare young drivers with older. 

One point of concern about these tests is that since the data come from one state (Iowa), the results are only applicable to car crashes there and cannot be generalized much further. If other Midwestern states collected similar data, then tests like those conducted in this study and those proposed above could be run again to gather more comprehensive data about the relationship between driving conditions and crash severity in the whole region. 


\newpage 

## References 

Ehsani, Johnathan P. (2016, June 24). "Teen Drivers' Perceptions of their Peer Passengers." Retrieved from <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4920469/> 

FHWA. (2018, February 17). "How Do Weather Events Impact Roads?" Retrieved from <https://ops.fhwa.dot.gov/weather/q1_roadimpact.htm>

Governor's Highway Safety Association (GHSA). (2019). "Seat Belt Laws." Retrieved from  <https://www.ghsa.org/state-laws/issues/Seat-Belts> 

Iowa Department of Transportation (DOT). (2017). "Crash Person Data Location." Retrieved from <https://catalog.data.gov/dataset/crash-person-data-location> 

Marmor, M. and Marmor, N. E. (2006, May). "Slippery Road Conditions and Fatal Motor Vehicle Crashes in the Northeastern United States, 1998-2002." Retrieved from <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1470576/>

United States Department of Transportation. (2016). National Highway Safety Traffic Administration. Retrieved from <https://www.nhtsa.gov/risky-driving/seat-belts> 

\newpage 

## Appendix 

### Randomization Test Code 

#### Data Summary 
```{r ref.label='rand data', echo = T, eval = F}
```

#### Randomization 
```{r ref.label='ran randomization', echo = T, eval = F}
```

#### Distribution and Assumptions
```{r ref.label='rand dist', echo = T, eval = T}
```

#### P-Value Calculation
```{r ref.label='rand p', echo = T, eval = F}
```


### T-Test Code 

#### Data Summary 
```{r ref.label='t data', echo = T, eval = F}
```

#### Check for Assumptions
```{r ref.label='t assumptions', echo = T, eval = F}
```

#### T-Test and P-Value
```{r ref.label='t statistic', echo = T, eval = F}
```


### Chi Square Test Code 

#### Data Summary 

#### Chi-Square Test 
```{r ref.label='chi', echo = T, eval = F}
```

#### Check for Assumptions 
```{r ref.label='exp', echo = T, eval = F}
```



